---
title: "Xg Model"
output: pdf_document
date: "2025-04-19"
---

# Overview

Football is a sport where every year something new is introduced, so in the year 2012 a new metric X-g was introduced 
that changed the whole game and how we look at it. Let us first understand what X-g is & how it works

X-g is a statistic metric that is used in football to predict the probability of a shot being a goal. It assigns
a value of 0-1 to each shot where if assigned 0.2, 2 out of 10 shots will be a goal meaning higher the value higher 
is the chance of being a goal.

I will be training a X-g Model that can predict the X-g value of a shot being a goal or not based on historical data of similar shots

# Gathering & Cleaning Data

Loading the libraries & the Data set

```{r, warning=FALSE,message=FALSE}
.libPaths("/home/rgds/R/x86_64-pc-linux-gnu-library/4.4")  #path to library

library(jsonlite)
library(tidyverse)
library(ggsoccer)
library(caret)
library(pROC)
```


The dataset I will be using contains the **Event Data** of the matches played in the **Bundesliga** (German First Div)
which i got from [wyscout](https://doi.org/10.6084/m9.figshare.7770599.v1).

```{r}
germany <- jsonlite::fromJSON("events/events_Germany.json")
germany <- as.data.frame(germany)         #converting it to a data frame
head(germany)
```

We see that there is alot of unwanted variables in this data so first need to clean it throughly & also
extract the x,y co-ordinates of the players who shot the ball for calculating new metrics.

```{r}
shots <- germany %>%
       
  filter(subEventName == "Shot") %>%                                            #filtering the rows with shot                                       
  mutate(goal = if_else(sapply(tags,function(x) any(x$id[1] == 101)),1,0)) %>%  #Goal column with 1 , 0  
  mutate(X = sapply(positions,function(x) X = x$x[1])) %>%                       #x co-rd of shot
  mutate(Y = sapply(positions,function(x) Y = x$y[1])) %>%                       #y co-rd of shot
  select(goal,X,Y) 

```

we could have just used the normal x-y Co-ordinate of the players, but to step things up I will be using the convention used by 
[FOT(friends of Tracking)](https://www.youtube.com/watch?v=bpjLyFyLlXs), who have made a similar project about X-g but in python.

So the two new metrics that i will be using are the **Distance** & **Shot Angle** metric as In Football, understanding the distance from the goal 
and the shooting angle against the goal post is crucial for maximizing scoring chances.

## Distance Helper function

The distance metric is the straight-line distance from the player's position to the center of the goal line.
it is calculated by using the pythagoreon theorem


```{r}
distance <- function(X,Y) {
                                       
  x <- 95                              # the width of the football field meters  
  y <- 75                              # the height of the football field in meters
  
  x_dist <- (100 - X)*x/100        
  y_dist <- abs(50 - Y)*y/100      
  
  distance <- sqrt(x_dist^2 + y_dist^2) #x-dist is the horizontal distance from the player to the goal line.
                                        #(100%) along the x-axis "right edge
                                        
                                        #y-dist is the vertical distance from the player to the center o
                                        #of the goal.(50%) along the y-axis "middle"
  
}
```

## Shot Angle Helper function

The goal angle metric helps us to determine how much of the goal post is actually visible to the player while shooting
meaning the bigger the goal angle better the chances of scoring.

```{r}

goal_angle <- function(X,Y) {
  
  x <- 95                              # the width of the football field meters  
  y <- 75                              # the height of the football field in meters
  
  x_dist <- (100 - X)*x/100            # the horizontal distance from the shot taken to the right edge of the goal line(x-axis 100%)
  y_dist <- abs(50 - Y)*y/100          # the vertical distance from the shot taken to the middle 
  
  
  angle <- atan((7.32*x_dist)/(x_dist^2 + y_dist^2 - (7.32/2)^2))  #the "7.32" is the width of the goal post
  angle <- ifelse(angle < 0,angle + pi,angle)                         
  
}
```


Now let's use these function to calculate the new metrics

```{r}
shots <- shots %>% 
              mutate(Distance = distance(X,Y)) %>%
              mutate(Shot_angle = goal_angle(X,Y)) %>%
              mutate(goal = as.factor(goal)) 


```


# EDA - Analysis before training the model

plotting the shot taken along with the ones that were counted as a goal.

```{r}
set.seed(123)  

sampled_shots <- shots[sample(nrow(shots), 1000), ]     #the data set is too big so sampling 1000 random rows


ggplot(sampled_shots) +
  annotate_pitch(colour = "black",fill = "white") + 
  geom_point(aes(x = X,y = Y,color = goal),alpha = 0.7) + #used gg soccer to plot the pitch
  coord_cartesian(xlim = c(45, 100))

```


with the visuals we can easily deduce how these two metric are very important for our X-g model:

- Distance =  we can see how most of the goals have been scored within or near the D-area meaning that the closer the player is higher 
  is the chance of scoring the goal.

- Shot Angle = the angel the shot was taken from can also be very important irrespective of the distance the ball was shot from , as 
  we see the front of the goal line has a concentration of goal compared to the left and right side of the goal post indicating the shot angle matters alot
  

# Model Training

Now coming to the main part of my X-G model project that is training the model to predict the Expected goals, There are plethora of ML algorithms that I can use  
to train my model but since I have already done a project on [Testing ML prediction algorithms](https://rpubs.com/Yvescandoit/m9999) , I'll be going with 
**Random Forest** that has the most accuracy among all the prediction algorithm.


```{r, warning=FALSE}

train_dat <- createDataPartition(shots$goal, p = 0.7,list = FALSE)  #splitting into 70% train and rest as test

train_data <- shots[train_dat,]
test_data <- shots[-train_dat,]
```

I will also be cross validating using k-fold CV 5 times and then use the model to predict the test data.

```{r}
set.seed(101)

trcon <- trainControl(method = "cv",number = 5)

train_rf <- train(goal ~.,data = train_data,method = "rf",trControl = trcon,verbose = FALSE)
```


# Accuracy of the Model (Model Fit)

Firstly, let's test how accurately the model fits on our testing data through a confusion matrix.

```{r}

prediction <- predict(train_rf,test_data)

confusionMatrix(prediction,test_data$goal)
```

The **Accuracy** of our model is *88%* which is really good but we see this model is heavily skewed towards predicting non-goal(0) 
which is due to class imbalance, meaning that the high accuracy of this model is misleading as it only predicts non-goals(0)

we can also see this in the Predictive values as well where it has a *high PPV of 90%* for non goals while its *NPV is only 36%* for goals 
meaning it fails to reliably predict actual goal scoring opportunities. 

In sports analytics , this model would be nothing more than a model that predicts every shot as a non goal which 
can be fatal for tactical decision-making about scoring potential in a match.


# ROC curve 

A Receiver operating characteristic curve is one of the best ways to asses model fit for models that work on binary, which perfectly fits our test case.
It works by plotting TPR(true positive rate) on y-axis & FPR(false positive rate) on x-axis at different threshold(eg changing what score counts as a positive
prediction).

A perfect model would shoot up to the left top edge(100% correct prediction) whereas a diagonal line means a random guesser(like a coin flip)
so anything between 0.6-0.9 would be a good model while 1 would be the perfect model.


```{r}

pred_roc <- predict(train_rf,test_data,type = "prob")[,"1"] #only positive values

roc_data <- roc(test_data$goal ,pred_roc)          #predicted goal

ggroc(roc_data,colour = "green",size = 2) 

```



# Things we can improve 

The model we just trained had a lot of flaws that could lead to a disaster if used in a real setting, so here are some of the things
that we can improve on

- Data = The main reason for our model to be this heavily skewed was the class imbalance of the data set where the the data for non-goal
  overpowered the data with goal, so opting for a data set with similar class balance would not have lead to this much of a accuracy paradox where it
  only predicts the majority class
  
- Metric = The metrics that we used to train the model is the type of data that can easily be collected(X,Y co-rd of player shot) or can be calculated(Goal angle,
  Distance) , but metrics like *no of defenders, position of the keeper, the body part used to shoot* are the type of data hard to collect, so using complex
  metrics like these would have definitely improved our models predictive abilities.
  
# conclusion

In this project we cleaned, explored the data while also creating new metrics that was used to train our model, after checking the accuracy of the model
which was not up to the mark we also came up with ways that can improve our model's predictive abilities.

In conclusion, as discussed earlier in the overview **Football is a sport which keeps on evolving & so does the data collection of it** so in the near future
there will be a time where we can implement data in our X-G model that we now consider impossible to collect.

## MIT License

Copyright (c) 2025 XgModel authors

Permission is hereby granted, free of charge, to any person obtaining a copy
of this software and associated documentation files (the "Software"), to deal
in the Software without restriction, including without limitation the rights
to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
copies of the Software, and to permit persons to whom the Software is
furnished to do so, subject to the following conditions:

The above copyright notice and this permission notice shall be included in all
copies or substantial portions of the Software.

THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
SOFTWARE.
